---
title: "CTData Exercise"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
# Set global options to hide messages/warnings but show code and output
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Load packages
```{r}
library(tidyverse)  # Loads the tidyverse collection of packages for data manipulation and visualization
library(psych)      # Loads the psych package for descriptive statistics and data analysis
```
# Read in Data
```{r}
data <- read.csv("CTData_Research-Analyst_Technical-Exercise_DATA.csv")
```

# Data Processing Steps & Analyses
## Descriptive analyses
```{r}
describe(data)  # Provides descriptive statistics for all variables in the dataset
unique(data$SMI)  # Lists unique values in the SMI column
```

# What percentage of respondents were currently using non-parental care for any child at the time of the survey? Include the sample size for your analysis.
```{r}
# Counting the number of "1" responses for each variable for using non-parental care
data$yes_count <- rowSums(data[, c("CT.POLICY.002.a", "CT.POLICY.003.a", "CT.POLICY.004.a", 
                                   "CT.POLICY.005.a", "CT.POLICY.006.a")] == 1, na.rm = TRUE)

# Calculate the number of "yes" responses for ct.policy.001
total_yes_ct_policy_001 <- sum(data$CT.POLICY.001 == 1, na.rm = TRUE)

# Calculate the number of respondents using non-parental care for any child 
using_non_parental_care <- ifelse(data$yes_count > 0, 1, 0)

n_using_non_parental_care <- sum(using_non_parental_care, na.rm = TRUE)

# Calculate the percentage of respondents using non-parental care (denominator is total_yes_ct_policy_001)
percentage_using_non_parental_care <- (n_using_non_parental_care / total_yes_ct_policy_001) * 100

# Output the percentage and sample size (total_yes_ct_policy_001)
cat("Percentage of respondents using non-parental care: ", percentage_using_non_parental_care, "%\n")
cat("Total number of respondents using non-parental care for any child: ", n_using_non_parental_care)
cat("Total number of respondents", total_yes_ct_policy_001)
```

# What percentage of young children were currently receiving non-parental care at the time of the survey? Include the sample size for your analysis. 
```{r}
# Count the number of children under 6 years old reported by respondents
total_children_under_6 <- sum(data$CT.DEMO.005, na.rm = TRUE)

# Count the number of children receiving non-parental care
# This sums "yes" responses (1) for all applicable children
children_receiving_care <- sum(
  rowSums(data[, c("CT.POLICY.002.a", "CT.POLICY.003.a", "CT.POLICY.004.a", 
                   "CT.POLICY.005.a", "CT.POLICY.006.a")] == 1, na.rm = TRUE)
)

# Calculate the percentage of young children receiving non-parental care
percentage_receiving_care <- (children_receiving_care / total_children_under_6) * 100

# Output the percentage and sample size
cat("Percentage of young children receiving non-parental care: ", percentage_receiving_care, "%\n")
cat("Total number of children receiving non-parental care: ", children_receiving_care, "\n")
cat("Total number of children under 6 years old: ", total_children_under_6, "\n")
```
# What percentage of respondents within each household income category (relative to state median income) were currently using non-parental care for any child? Please include the sample size for your analysis and a data visualization.
```{r}
# Convert blank strings to NA in the SMI column
data$SMI <- na_if(data$SMI, "")

# Create a new column for using_non_parental_care
data$using_non_parental_care <- ifelse(data$yes_count > 0, 1, 0)

# Group data by SMI and calculate the necessary values
income_analysis <- data %>%
  group_by(SMI) %>%
  summarise(
    total_respondents = sum(CT.POLICY.001 == 1, na.rm = TRUE),
    n_using_non_parental_care = sum(using_non_parental_care == 1, na.rm = TRUE),
    percentage_using_non_parental_care = (n_using_non_parental_care / total_respondents) * 100,
    .groups = "drop"  # Drop grouping after summarizing
  ) %>%
  filter(!is.na(SMI))  # Exclude NA groups

print(income_analysis)

# Calculate the totals
total_row <- data.frame(
  SMI = "Total",
  total_respondents = sum(income_analysis$total_respondents, na.rm = TRUE),
  n_using_non_parental_care = sum(income_analysis$n_using_non_parental_care, na.rm = TRUE),
  percentage_using_non_parental_care = (sum(income_analysis$n_using_non_parental_care, na.rm = TRUE) / 
                                          sum(income_analysis$total_respondents, na.rm = TRUE)) * 100
)

# Combine the summary and total row
income_analysis <- bind_rows(income_analysis, total_row)

# Print the updated income analysis table
print(income_analysis)
```

## Create graph
```{r}
# Create a bar plot for visualization excluding the total row
visualization_data <- income_analysis %>% filter(SMI != "Total")  # Exclude the total row

# Define the order of SMI categories
visualization_data$SMI <- factor(visualization_data$SMI, 
                                  levels = c("Below 30% of SMI", 
                                             "Between 30-60% of SMI", 
                                             "Between 60-85% of SMI", 
                                             "Above 85% of SMI"))

ggplot(visualization_data, aes(x = SMI, y = percentage_using_non_parental_care)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Format y-axis as percentage
  labs(
    title = "Percentage of Respondents Using Non-Parental Care by Income Category",
    x = "Income Category (Relative to State Median Income)",
    y = "Percentage Using Non-Parental Care"
  ) +
  geom_text(aes(label = paste0(round(percentage_using_non_parental_care, 1), "%")), 
            vjust = -0.5, size = 3) +  # Add percentage labels on the bars
  theme_minimal()
```

# Is the variation among household income groups in the likelihood of using nonparental care for any child statistically significant? Report the test(s) you used, the value(s) of the test statistic(s), the degrees of freedom, and the p-value(s). Describe this result and what it means in language that is accessible to a general audience. 

## Data cleaning
```{r}
# Check unique values in the SMI column after trimming whitespace
unique_values <- unique(trimws(data$SMI))
print(unique_values)

# Clean the SMI values and filter out NA or blank entries
cleaned_data <- data %>%
  filter(!is.na(SMI) & trimws(SMI) != "")

# Check for any non-standard entries in the cleaned dataset
unique_cleaned_values <- unique(trimws(cleaned_data$SMI))
print(unique_cleaned_values)

# Check for non-visible characters
cleaned_data <- cleaned_data %>%
  filter(SMI != " " & SMI != "")  # Filter out spaces and empty strings

# Double-check unique values again
unique_cleaned_values <- unique(trimws(cleaned_data$SMI))
print(unique_cleaned_values)

cleaned_data$SMI <- as.factor(cleaned_data$SMI)
cleaned_data$using_non_parental_care <- as.factor(cleaned_data$using_non_parental_care)
```

## Statistical analysis
```{r}
# Create a contingency table for the number of respondents using non-parental care
contingency_table <- table(cleaned_data$SMI, cleaned_data$using_non_parental_care)
print(contingency_table)

# Perform the Chi-Square test
chi_square_test <- chisq.test(contingency_table)

# Extracting the results
chi_square_statistic <- chi_square_test$statistic
degrees_of_freedom <- chi_square_test$parameter
p_value <- chi_square_test$p.value

# Print the results
cat("Chi-Square Statistic:", chi_square_statistic, "\n")
cat("Degrees of Freedom:", degrees_of_freedom, "\n")
cat("P-Value:", p_value, "\n")
```
